Type-Safe Modular Parsing

### 2017-01-12 paper/sections/Discussion.tex
Revised 5.2

Having pointed out that our goal is to realize modular parsers, we expect the code written in an extensible way. We have been focusing on how to extend a language with some additional cases, just as many extensible compilers do, nevertheless, in other cases, languages are defined separately, and the composition of them is again a matter of code reuse. In this section, we propose a framework for defining small languages as modular components, in order for better composition.

The framework we use for a language, or rather a language component in our case study, consists of three parts: syntax, parsing and operations on the AST. More
specifically:
\begin{itemize}
    \item \textbf{Object Algebra interface:} defined as a trait for the abstract syntax. The type parameters represent multiple sorts of syntax, and the methods are constructs.
\item \textbf{Parser:} implementation of the parser with lexing, using inheritance and Object Algebras, with its scoping trait abstracted over the type parameters.
\item \textbf{Algebras (optional):} concrete operations on ASTs, such as pretty-printing.
\end{itemize}

Taking again the example in Section~\ref{subsec:parsingwithoa}. The language constructs consists of literals, additions and variables. First we define the
language component in a Scala object, shown in the left part of Figure~\ref{fig:lng-components}.

@@

For the example in Section~\ref{subsec:differentsyntax}, instead of building inheritance directly, we define the language separately,
which consists of types and typed lambda expressions. Shape of the code is presented in the right of Figure~\ref{fig:lng-components}.

At this point, one may want those two language components to be composed into one, and meanwhile ensure that the one is still a modular
component ready for future composition. In that case modularity is realized over higher-order hierarchies. The composed language \lstinline{VarLamExpr}
is shown below:

\lstinputlisting[linerange=82-91]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
Note that lexing code is automatically combined under multiple inheritance. The combined language shows perfect modularity. The only
drawback is that the glue code of composition appears to be boilerplate. As shown above, we are combining ASTs, parsers and pretty printers of
\lstinline{VarExpr} and \lstinline{TypedLam} respectively. Such a pattern refers to \textit{family polymorphism}[]. Unfortunately Scala does not
support this feature, nonetheless, one can avoid such boilerplate by using metaprogramming techniques.

\subsection{Alternative Techniques}
Under our modular parsing framework, we use Packrat parsing as the underlying parsing technique, OO inheritance with method overriding for composing and extending parsers, and Object Algebras for parsing extensible ASTs. However, our framework is more general and thus more powerful, because those aspects are orthogonal to each other, and hence can have their alternatives.

>>>>

Modular parsing not only enables us to build a corresponding parser which evolves with the language together, but also allows us to abstract language features as reusable, independent components. Generally, a language feature includes related abstract syntax, methods to \textit{build} the syntax (parsing), and methods to \textit{process} the syntax (evaluation, pretty-printing, etc.). From this perspective, not only one language, but many languages can be developed in a modular way, with common language features reused.

Instead of design and build a language from scratch, we can easily add a new feature by reusing the corresponding language component. For example, if a language is composed with a component of boolean expressions, including if-then-else, it immediately knows how to parse, traverse, and pretty-print if-then-else structure! It could be very useful for rapid developing DSLs.

For implementation, a language component is represented by a Scala object, and it consists of three parts: Object Algebra interface, parser, and Object Algebras.

\begin{itemize}
    \item \textbf{Object Algebra interface:} defined as a trait for the abstract syntax. The type parameters represent multiple sorts of syntax, and the methods are constructs.
    \item \textbf{Parser:} corresponding parser of the abstract syntax, written in a modular way as we demonstrated in previous sections.
    \item \textbf{Object Algebras (optional):} concrete operations on ASTs, such as pretty-printing.
\end{itemize}

We take the example in Section~\ref{subsec:parsingwithoa} again, which is a language of literals, additions and variables. It is defined as a
language component on the left side of Figure~\ref{fig:lng-components}.
Then for the extension of types and lambda abstractions in Section~\ref{subsec:differentsyntax}, instead of inheriting from the previous language directly, we define it as another independent language component. The code is presented on the right side of Figure~\ref{fig:lng-components}.

@@

As shown below, those two language components can be merged into one to obtain the language we want. Futhermore, the new language is still a modular
component ready for future composition. In that case modularity is realized over higher-order hierarchies.

\lstinputlisting[linerange=82-91]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
Such composition shows perfect modularity. The only
drawback is that the glue code of composition appears to be boilerplate. As shown above, we are combining ASTs, parsers and pretty printers of
\lstinline{VarExpr} and \lstinline{TypedLam} respectively. Such a pattern refers to \textit{family polymorphism}[]. Unfortunately Scala does not
support this feature, nonetheless, one can avoid such boilerplate by using metaprogramming techniques.

\subsection{Alternative Techniques}

Under our modular parsing framework, we use Packrat parsing as the underlying parsing technique, OO inheritance with method overriding for composing and extending parsers, and Object Algebras for parsing extensible ASTs. However, our framework is more general and thus more powerful, because those aspects are orthogonal to each other, and hence can have their alternatives.
